Make any stupid plot you want

### 2021-04-16 draft/sections/intro.tex
data->monoid sections + reference renaming

\documentclass[../main.tex]{subfiles}
\begin{document}
\section{Introduction}
\label{sec:intro}
\begin{figure}[h]
    \includegraphics[width=1\textwidth]{figures/intro/functions.png}
    \caption{Visualizations are made up of transformations from data into visual representation. These functions transform individual data values to visual representation, such as date to x position or latitude to y position. These functions are composed into the assembly of all these transformations into a visual mark, such as a line or point. %%% break this out into seperate sentences to break 
    The same variable can be mapped in different ways, for example line is mapped to a color in the scatter plot and to y position in the line plot.}
    \label{fig:intro_artist}
\end{figure}
Visualization is the transformation of data into visual representation. As illustrated by \autoref{fig:intro_artist}, these translations are both at the level of the individual variable and the entire record. In the case of the scatter plot, the latitude and longitude are encoded as the x and y position, respectively, while the temperature and station are represented by the face and edge colors. A row in the table is collectively encoded as a point mark. None of these encodings are fixed, as evidenced by temperature being translated into the y value in the case of the line plot. The station is now the source of the color of the entire line, and the date is the x position. As with scatter, the encodings of the individual transformations, which again are on values from the same record in the table, are composited into a line mark. It is these raw transformations from data space to visualization space that are implemented by building block level visualization libraries, named as such because the functions provided by the library can be composited in any number of ways to yield visualizations \cite{wongsuphasawatNavigatingWideWorld2021}. We propose that like physical building blocks, building block libraries must provide a collection of well defined pieces that can be composed in whichever ways the blocks fit together. We specify that a valid visualization block is a structure preserving transformation from data to visual space, and we define structure in terms of \textit{continuity} and \textit{equivariance}. We then use this model to develop a design specification for the components of a building block visualization library. The notion of self contained, inherently modular, building blocks lends itself naturally to a functional paradigm of visualization \cite{hughesWhyFunctionalProgramming1989}. We adopt a functional model for a redesign because the lack of side effects means functional architecture can be evaluated for correctness, functional programs tend to be shorter and clearer, and are well suited to distributed, concurrent, and on demand tasks\cite{huHowFunctionalProgramming2015}.

@@

\subsection{Tools}
\label{sec:intro_data_tools}

@@

\begin{figure}[H]
    \includegraphics[width=1\textwidth]{figures/intro/dataset_diagram.png}
    \caption{One way to describe data is by the connectivity of the points in the dataset. A database for example is often discrete unconnected points, while an image is an implicitely connected 2D grid. This image is from the Data Representation chapter of the MayaVi 4.7.2 documentation.\cite{DataRepresentationMayavi}}
    \label{fig:intro_data_format}
\end{figure}
Many building block libraries carry multiple models of data internally because they cannot assume a data structure. Algorithms are designed such that the structure of data is assumed, as described in Tory and Möller's taxonomy \cite{ToryRethinkingVisualization2004}, and by definition building block libraries try to  provide the components to build any sort of visualization. Matplotlib, D3\cite{bostockDataDrivenDocuments2011}, and VTK \cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk} and its derivatives such as MayaVi\cite{RamachandranMayaVI2011} and extensions such as ParaView\cite{ahrens2005paraview} and the infoviz themed Titan\cite{brianwylieUnifiedToolkitInformation2009}. Where GoG and ImageJ type libraries have coherant APIs for their visualization tools because the data structure is the same, the APIs for visualizations in Matplotlib, D3, and VTK are significantly dependent on the structure of the data it expects. VTK has codified this in terms of \textit{continuity} based data representations, as illustrated in figure~\ref{fig:intro_data_format}. This API choice can lead to visualizations that break \textit{continuity} when fed into visualizations with different assumptions about structure. The lack of consistent data model can also mean no consistent way of updating the data and therefore no way of guaranteeing that the views are in sync, in visualizations that consistent of multiple views of the same datasource, such as dashboards\cite{a.sarikayaWhatWeTalk2019,fewDashboardConfusionRevisited2007}. To resolve this issue, our functional model takes as input a structure aware data abstraction general enough to provide a common interface for many different types of visualization.

@@

\begin{figure}[H]
    \includegraphics[width=1\textwidth]{figures/intro/munzner_datatypes.png}
    \caption{Values in a dataset have keys associated with them that describe where the value is in the dataset. These keys can be indexers or semantically meaningful;for example, in a table the keys are the variable name and the row ID. In the data cube,  the keys is the row, column, and cell ID, and in the map the key is the position in the grid. Image is figure 2.8 in Munzner's Visualization Analysis and Design\cite{munznerVisualizationAnalysisDesign2014}}
    \label{fig:intro:keys_values}
\end{figure}

The \textit{continuity} can often be described by some variables in the dataset; this is formalized Munzner's notion of metadata as \textit{keys} into the data structure that return associated \textit{values}\cite{munznerChDataAbstraction}. As shown in \autoref{fig:intro:keys_values}, keys can be labeled indexes, such as the attribute name and row ID, or physical entities such as locations on a map. We propose that information rich metadata are part of the components and instead the values are keyed on coordinate free structural ids. In contrast to Munzner's model where the semantic meaning of the key is tightly coupled to the position of the value in the dataset, our model considers keys to be a pure reference to topology. This allows the metadata to be altered, without imposing new semantics on the underlying structure, for example by changing the coordinate systems or time resolution. This value agnostic model also supports encoding datasets where there may be multiple independent variables - such as a measure of plant growth given variations in water, sunlight, and time - without having to assume any one variable is inducing the change in growth. For building block library developers, this means the components are able to fully traverse the data structures without having to know anything about the values or the semantic meaning of the structure. Since these components are by design \textit{equivariant} and \textit{continuity} preserving, domain specific library developers in different domains that both rely on the same continuity, for example 2D continuity, can then safely reuse the components to build tools that can safely make domain specific assumptions.

\subsection{Contribution}
The contribution of this work is 
\begin{enumerate}
  \item formalization of the topology preserving relationship between data and graphic via continuous maps \autoref{sec:graphic_base}
  \item formalization of property preservation from data component to visual representation as monoid action equivariant maps \autoref{sec:artist_nu}
  \item functional oriented visualization architecture built on the mathematical model to demonstrate the utility of the model \autoref{sec:artist_q}
  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model. \autoref{sec:code_artist}
\end{enumerate}
\end{document}

>>>>

\documentclass[../main.tex]{subfiles}
\begin{document}
\section{Introduction}
\label{sec:intro}
\begin{figure}[h]
    \includegraphics[width=1\textwidth]{figures/intro/functions.png}
    \caption{Visualizations are made up of transformations from data into visual representation. These functions transform individual data values to visual representation, such as date to x position or latitude to y position. These functions are composed into the assembly of all these transformations into a visual mark, such as a line or point. %%% break this out into seperate sentences to break 
    The same variable can be mapped in different ways, for example line is mapped to a color in the scatter plot and to y position in the line plot.}
    \label{fig:intro:artist}
\end{figure}
Visualization is the transformation of data into visual representation. As illustrated by \autoref{fig:intro:artist}, these translations are both at the level of the individual variable and the entire record. In the case of the scatter plot, the latitude and longitude are encoded as the x and y position, respectively, while the temperature and station are represented by the face and edge colors. A row in the table is collectively encoded as a point mark. None of these encodings are fixed, as evidenced by temperature being translated into the y value in the case of the line plot. The station is now the source of the color of the entire line, and the date is the x position. As with scatter, the encodings of the individual transformations, which again are on values from the same record in the table, are composited into a line mark. It is these raw transformations from data space to visualization space that are implemented by building block level visualization libraries, named as such because the functions provided by the library can be composited in any number of ways to yield visualizations \cite{wongsuphasawatNavigatingWideWorld2021}. We propose that like physical building blocks, building block libraries must provide a collection of well defined pieces that can be composed in whichever ways the blocks fit together. We specify that a valid visualization block is a structure preserving transformation from data to visual space, and we define structure in terms of \textit{continuity} and \textit{equivariance}. We then use this model to develop a design specification for the components of a building block visualization library. The notion of self contained, inherently modular, building blocks lends itself naturally to a functional paradigm of visualization \cite{hughesWhyFunctionalProgramming1989}. We adopt a functional model for a redesign because the lack of side effects means functional architecture can be evaluated for correctness, functional programs tend to be shorter and clearer, and are well suited to distributed, concurrent, and on demand tasks\cite{huHowFunctionalProgramming2015}.

@@

\subsection{Tools}
\label{sec:intro:data:tools}

@@

\begin{figure}[H]
    \includegraphics[width=1\textwidth]{figures/intro/dataset_diagram.png}
    \caption{One way to describe data is by the connectivity of the points in the dataset. A database for example is often discrete unconnected points, while an image is an implicitely connected 2D grid. This image is from the Data Representation chapter of the MayaVi 4.7.2 documentation.\cite{DataRepresentationMayavi}}
    \label{fig:intro:data:format}
\end{figure}
Many building block libraries carry multiple models of data internally because they cannot assume a data structure. Algorithms are designed such that the structure of data is assumed, as described in Tory and Möller's taxonomy \cite{ToryRethinkingVisualization2004}, and by definition building block libraries try to  provide the components to build any sort of visualization. Matplotlib, D3\cite{bostockDataDrivenDocuments2011}, and VTK \cite{hanwellVisualizationToolkitVTK2015,geveci2012vtk} and its derivatives such as MayaVi\cite{RamachandranMayaVI2011} and extensions such as ParaView\cite{ahrens2005paraview} and the infoviz themed Titan\cite{brianwylieUnifiedToolkitInformation2009}. Where GoG and ImageJ type libraries have coherant APIs for their visualization tools because the data structure is the same, the APIs for visualizations in Matplotlib, D3, and VTK are significantly dependent on the structure of the data it expects. VTK has codified this in terms of \textit{continuity} based data representations, as illustrated in figure~\ref{fig:intro:data:format}. This API choice can lead to visualizations that break \textit{continuity} when fed into visualizations with different assumptions about structure. The lack of consistent data model can also mean no consistent way of updating the data and therefore no way of guaranteeing that the views are in sync, in visualizations that consistent of multiple views of the same datasource, such as dashboards\cite{a.sarikayaWhatWeTalk2019,fewDashboardConfusionRevisited2007}. To resolve this issue, our functional model takes as input a structure aware data abstraction general enough to provide a common interface for many different types of visualization.

@@

\begin{figure}[H]
    \includegraphics[width=1\textwidth]{figures/intro/munzner_datatypes.png}
    \caption{Values in a dataset have keys associated with them that describe where the value is in the dataset. These keys can be indexers or semantically meaningful;for example, in a table the keys are the variable name and the row ID. In the data cube,  the keys is the row, column, and cell ID, and in the map the key is the position in the grid. Image is figure 2.8 in Munzner's Visualization Analysis and Design\cite{munznerVisualizationAnalysisDesign2014}}
    \label{fig:intro:keys:values}
\end{figure}

The \textit{continuity} can often be described by some variables in the dataset; this is formalized Munzner's notion of metadata as \textit{keys} into the data structure that return associated \textit{values}\cite{munznerChDataAbstraction}. As shown in \autoref{fig:intro:keys:values}, keys can be labeled indexes, such as the attribute name and row ID, or physical entities such as locations on a map. We propose that information rich metadata are part of the components and instead the values are keyed on coordinate free structural ids. In contrast to Munzner's model where the semantic meaning of the key is tightly coupled to the position of the value in the dataset, our model considers keys to be a pure reference to topology. This allows the metadata to be altered, without imposing new semantics on the underlying structure, for example by changing the coordinate systems or time resolution. This value agnostic model also supports encoding datasets where there may be multiple independent variables - such as a measure of plant growth given variations in water, sunlight, and time - without having to assume any one variable is inducing the change in growth. For building block library developers, this means the components are able to fully traverse the data structures without having to know anything about the values or the semantic meaning of the structure. Since these components are by design \textit{equivariant} and \textit{continuity} preserving, domain specific library developers in different domains that both rely on the same continuity, for example 2D continuity, can then safely reuse the components to build tools that can safely make domain specific assumptions.

\subsection{Contribution}
The contribution of this work is 
\begin{enumerate}
  \item formalization of the topology preserving relationship between data and graphic via continuous maps \autoref{sec:math:graphic:base}
  \item formalization of property preservation from data component to visual representation as monoid action equivariant maps \autoref{sec:math:artist:nu}
  \item functional oriented visualization architecture built on the mathematical model to demonstrate the utility of the model \autoref{sec:math:artist:q}
  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model. \autoref{sec:math:code:artist}
\end{enumerate}
\end{document}

### 2021-04-16 draft/sections/math.tex
data->monoid sections + reference renaming

\section{Topological Artist Model}
\label{sec:tam}
To guide the implementation of structure preserving building block components, we develop a mathematical formalism of visualization that specifies how these components preserve \textit{continuity} and \textit{equivariance}. Inspired by the somewhat analogous component in Matplotlib\cite{hunterArchitectureOpenSource}, we call the transformation from data space to graphic that these building block components implement the \textit{artist}.
\begin{equation}
    \label{eq:artist}
    \mathscr{\vartist}: \mathscr{\dtotal} \rightarrow \mathscr{\gtotal}
\end{equation}
The \textit{artist} $\mathscr{\vartist}$ is a map from the data $\mathscr{\dtotal}$ to graphic $\mathscr{\gtotal}$ fiber bundles. To explain how the \textit{artist} is a structure preserving map from data to graphic, we first describe how we model data (\ref{sec:data}) and graphics (\ref{sec:graphic}) as topological structures that encapsulate component types and continuity. We then discuss the maps from graphic to data (\ref{sec:graphic_base}, data components to visual components (\ref{sec:artist_nu}), and visual components into graphic (\ref{sec:artist_q}) that make up the artist.

\subsection{Data Space \dtotal}
\label{sec:data}
Building on Butler's proposal of using fiber bundles as a common data representation structure for visualization data\cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989}, a fiber bundle is a tuple $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$ defined by the projection map $\pi$
\begin{equation}
    \label{eq:fiber_bundle}
    \begin{tikzcd}
        \dfiber \arrow[r, hook] & \dtotal \arrow[r, "\pi"] & \dbase
    \end{tikzcd}
\end{equation}
that binds the components of the data in \dfiber\ to the continuity represented in \dbase. The fiber bundle models the properties of data component types \dfiber\ (\ref{sec:data_fiber}), the continuity of records \dbase\ (\ref{sec:data_base}), the collections of records \dsection\ (\ref{sec:data_section}), and the space \dtotal\ of all possible datasets with these components and continuity.

By definition fiber bundles are locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, meaning that over a localized neighborhood we can dispense with extra structure on \dtotal\ and focus on the components and continuity. We use fiber bundles as the data model because they are inclusive enough to express all the types of data described in section~\ref{sec:intro_data}.

\subsubsection{Variables in Fiber Space \dfiber}
\label{sec:data_fiber}
To formalize the structure of the data components, we use notation introduced by Spivak \cite{spivakSIMPLICIALDATABASES} that binds the components of the fiber to variable names. This allows us to describe the components in a schema like way. Spivak constructs a set \ftotal\ that is the disjoint union of all possible objects of types $\{\ftype_0, \ldots, \ftype_m\} \in \ftypes$, where \ftypes\ are the data types of the variables in the dataset. He then defines the single variable set \fttype\ 
\begin{equation}
    \label{eq:data_types}
    \begin{tikzcd}
        \fttype \arrow[r] \arrow[d, "\pi_{\fsection}"'] & \ftotal \arrow[d, "\pi"] \\
        \fnames \arrow[r, "\fsection"']                          & \ftypes       
    \end{tikzcd}
\end{equation}
which is \ftotal\ restricted to objects of type \ftype\ bound to variable name \fname. The \fttype\ lookup is by name to specify that every component is distinct, since multiple components can have the same type \ftype. Given \fsection, the fiber for a one variable dataset is
\begin{equation}
    \dfiber = \ftotal_{\fsection(\fname)} = \ftotal_{\ftype} 
\end{equation}
where \fsection\ is the schema binding variable name \fname\ to its datatype \ftype. A dataset with multiple variables has a fiber that is the cartesian cross product of $\ftotal_{\fsection}$ applied to all the columns:
\begin{equation}
F = \ftotal_{\fsection(\fname_{1})}\times \ldots \ftotal_{\fsection(\fname_{i})} \ldots\times \ftotal_{\fsection(\fname_{n})}
\end{equation}
which is equivalent to 
\begin{equation}
    \dfiber= \dfiber_{0} \times \ldots \times \dfiber_{i}\times\ldots\times \dfiber_{n}
\end{equation}
which allows us to decouple \dfiber\ into components $\dfiber_i$.

\begin{figure}[H]
    \includegraphics[width=1\textwidth]{figures/math/fiber.png}
    \caption{These two datasets have the same continuity, but different components and therefore different fibers. The 2D fiber \(\dfiber=\reals\times\reals\) encodes a dataset consisting of \textit{time} and \textit{temperature} components.  One dimension of the fiber encodes the range of possible values for the time component of the dataset, which is a subset of the \reals, while the other dimension encodes the range of possible values  \reals\ for the temperature component. The 3D fiber encodes a dataset consisting of\textit{temperature}, \textit{latitude}, and \textit{longitude} components.}
    \label{fig:data_fiber_example}
\end{figure}

For example, the records in the 2D fiber in \autoref{fig:data_fiber_example} are a pair of times and \textdegree K temperature measurements taken at those times. Time is a positive number of type \texttt{datetime} which can be resolved to floats $\ftotal_{\texttt{datetime}}= \reals$. Temperature values are real positive numbers $\ftotal_{\texttt{float}} = \realsp$. The fiber is 
\begin{equation*}
    \dfiber =  \reals \times \realsp 
\end{equation*} 
where the first component $F_0$ is the set of values specified by $(\fname=time,\, \ftype=\texttt{datetime},\, \fttype=\reals)$ and $F_1$ is specified by $(\fname=temperature,\, \ftype=\texttt{float},\, \fttype=\reals)$ and is the set of values $\fttype=\reals$. In the 3D fiber in \autoref{fig:fiber_example_cube}, time is replaced with location. This location variable is of type \texttt{point} and has two components \textit{latitude} and \textit{longitude} $\{(lat,lon) \in \reals^{2} \mid  -90\leq lat \leq 90,\, 0 \leq lon \leq 360\}$. The fiber for this dataset is
\begin{equation*}
    \dfiber = \reals \times \reals^{2} = \reals \times \reals \times \reals
\end{equation*} 
where the dimensionality of the fiber does not change, but the components of the fiber can be coupled. For example, \textit{location} can can either be specified as \((\fname = location,\, \ftype = \texttt{point},\, \fttype=\reals^{2})\)or \((\fname=latitude,\, \ftype=\texttt{float}, \, \fttype=\reals)\) and \((\fname=longitude,\, \ftype=\texttt{float},\,\fttype=\reals)\).

@@

\subsubsection{Measurement Scales: Monoid Actions}
\label{sec:data_monoid}
Implementing expressive visual encodings requires formally describing the structure on the components of the fiber, which we define by the actions of a monoid on the component. In doing so, we specify the properties of the component that must be preserved in a graphic representation. While structure on a set of values is often described algebraically as operations or through the actions of a group, for example Steven's scales \cite{stevensTheoryScalesMeasurement1946}, we generalize to monoids to support more component types. Monoids are also commonly found in functional programming because they specify compositions of transformations \cite{yorgeyMonoidsThemeVariations, stievenMonadJustMonoid2020}.

A monoid \cite{Monoid2021} $\monoid$ is a set with an associative binary operator $\ast:\monoid \times \monoid\rightarrow \monoid$. A monoid has an identity element $e\in \monoid$ such that $e\ast a= a \ast e = a$ for all $a \in \monoid$. As defined on a component of \dfiber, a left monoid action \cite{SemigroupAction2021,nlab:action} of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties:
\begin{align*}
    \textbf{associativity}\;& \text{for all } f,g \in \monoid_i \text{ and } x\in \dfiber_i,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
    \textbf{identity}\;& \text{for all } x\in \dfiber_i, e\in \monoid_i,\,  e\bullet x = x 
\end{align*}
As with the fiber \dfiber\, the total monoid space \monoid\ is the cartesian product
\begin{equation}
\monoid= \monoid_{0} \times \ldots \times \monoid_{i}\times \ldots \times\ldots \monoid_{n}
\end{equation}
of each monoid $\monoid_{i}$ on $\dfiber_{i}$.  The monoid is also added to the specification of the fiber $(\fname_i,\, \ftype_i,\, \fttype\, \monoid_i)$

Steven's described the measurement scales\cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale} in terms of the monoid actions on the measurements: nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}. For example, lets say one component of the data is weather states, such that the fiber component is $\dfiber_i=\{sun,\, clouds,\, rain,\, lightning,\, storm\}$. If the data is ordinal, that means some ordering action is applied to the set.

\begin{figure}[H]
\begin{subfigure}{.5\textwidth}
    \includegraphics[width=\textwidth]{figures/math/monoid_total.png}
    \caption{The identity action $\bullet$ maps the weather states to themselves.}
    \label{fig:monoid:total}
\end{subfigure}
\begin{subfigure}{.5\textwidth}
    \includegraphics[width=1\textwidth]{figures/math/monoid_partial.png}
    \caption{The action $\bullet$ is a monotonic map, meaning that it preserves the relative ordering despite mapping multiple elements to the same element.}
    \label{fig:monoid:partial}
\end{subfigure}
\end{figure}
\note{I think there's something about associativity here, but I'm not sure what}
 In \autoref{fig:monoid:partial}, the action $\bullet$ acts on the weather states such that the states that they are mapped into have the same relative ordering as the original states. Unlike group actions, the action $\bullet$ is not invertable and therefore a monoid action. We generalize to monoids specifically for the case of partially ordered scales are common in indicator rankings \cite{bruggemannRankingPrioritizationMultiindicator2011}.

>>>>

\section{Topological Artist Model}
\label{sec:tam}
To guide the implementation of structure preserving building block components, we develop a mathematical formalism of visualization that specifies how these components preserve \textit{continuity} and \textit{equivariance}. Inspired by the somewhat analogous component in Matplotlib\cite{hunterArchitectureOpenSource}, we call the transformation from data space to graphic that these building block components implement the \textit{artist}.
\begin{equation}
    \label{eq:artist}
    \mathscr{\vartist}: \mathscr{\dtotal} \rightarrow \mathscr{\gtotal}
\end{equation}
The \textit{artist} $\mathscr{\vartist}$ is a map from the data $\mathscr{\dtotal}$ to graphic $\mathscr{\gtotal}$ fiber bundles. To explain how the \textit{artist} is a structure preserving map from data to graphic, we first describe how we model data (\autoref{sec:math:data}) and graphics (\autoref{sec:math:graphic}) as topological structures that encapsulate component types and continuity. We then discuss the maps from graphic to data (\autoref{sec:math:graphic:base}, data components to visual components (\autoref{sec:math:artist:nu}), and visual components into graphic (\autoref{sec:math:artist:q}) that make up the artist.

\subsection{Data Space \dtotal}
\label{sec:math:data}
Building on Butler's proposal of using fiber bundles as a common data representation structure for visualization data\cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989}, a fiber bundle is a tuple $(\dtotal,\,\dbase,\,\pi ,\,\dfiber)$ defined by the projection map $\pi$
\begin{equation}
    \label{eq:fiber_bundle}
    \begin{tikzcd}
        \dfiber \arrow[r, hook] & \dtotal \arrow[r, "\pi"] & \dbase
    \end{tikzcd}
\end{equation}
that binds the components of the data in \dfiber\ to the continuity of the data encoded in \dbase. The fiber bundle models the properties of data component types \dfiber\ (\autoref{sec:math:data:fiber}), the continuity of records \dbase\ (\autoref{sec:math:data:base}), the collections of records \dsection\ (\autoref{sec:math:data:section}), and the space \dtotal\ of all possible datasets with these components and continuity. By definition fiber bundles are locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, meaning that over a localized neighborhood $U$ the total space is the cartesian product $\dbase\times\dfiber$. We use fiber bundles as the data model because they are inclusive enough to express all the types of structures of data described in \autoref{sec:intro:data:tools}

\subsubsection{Variables in Fiber Space \dfiber}
\label{sec:math:data:fiber}
To formalize the structure of the data components, we use notation introduced by Spivak \cite{spivakSIMPLICIALDATABASES} that binds the components of the fiber to variable names. This allows us to describe the components in a schema like way. Spivak constructs a set \ftotal\ that is the disjoint union of all possible objects of types $\{\ftype_0, \ldots, \ftype_m\} \in \ftypes$, where \ftypes\ are the data types of the variables in the dataset. He then defines the single variable set \fttype\ 
\begin{equation}
    \label{eq:data_types}
    \begin{tikzcd}
        \fttype \arrow[r] \arrow[d, "\pi_{\fsection}"'] & \ftotal \arrow[d, "\pi"] \\
        \fnames \arrow[r, "\fsection"']                          & \ftypes       
    \end{tikzcd}
\end{equation}
which is \ftotal\ restricted to objects of type \ftype\ bound to variable name \fname. The \fttype\ lookup is by name to specify that every component is distinct, since multiple components can have the same type \ftype. Given \fsection, the fiber for a one variable dataset is
\begin{equation}
    \dfiber = \ftotal_{\fsection(\fname)} = \ftotal_{\ftype} 
\end{equation}
where \fsection\ is the schema that binds a variable name \fname\ to its datatype \ftype. A dataset with multiple components has a fiber that is the cartesian cross product of $\ftotal_{\fsection}$ applied to all the columns:
\begin{equation}
F = \ftotal_{\fsection(\fname_{1})}\times \ldots \ftotal_{\fsection(\fname_{i})} \ldots\times \ftotal_{\fsection(\fname_{n})}
\end{equation}
which is equivalent to 
\begin{equation}
    \dfiber= \dfiber_{0} \times \ldots \times \dfiber_{i}\times\ldots\times \dfiber_{n}
\end{equation}
which allows us to decouple \dfiber\ into components $\dfiber_i$.

\begin{figure}[H]
    \includegraphics[width=1\textwidth]{figures/math/fiber.png}
    \caption{The fiber space is the set of all combinations of all theoretically possible values of the components.  The 2D fiber \(\dfiber=\reals\times\reals\) encodes the properties of \textit{time} and \textit{temperature} components.  One dimension of the fiber encodes the range of possible values for the time component of the dataset, which is a subset of the \reals, while the other dimension encodes the range of possible values  \reals\ for the temperature component. This means the fiber is the set of points \textit{(temperature, time)}that are all the combinations of \textit{temperatature} $\times$ \textit{time}. The 3D fiber encodes points at all possible combinations of \textit{temperature}, \textit{latitude}, and \textit{longitude}.}
    \label{fig:data:fiber:example}
\end{figure}

For example, the records in the 2D fiber in \autoref{fig:data:fiber:example} are a pair of \textit{times} and \textdegree K \textit{temperature} measurements taken at those times. Time is a positive number of type \texttt{datetime} which can be resolved to floats $\ftotal_{\texttt{datetime}}= \reals$. Temperature values are real positive numbers $\ftotal_{\texttt{float}} = \realsp$. The fiber is 
\begin{equation*}
    \dfiber =  \reals \times \realsp 
\end{equation*} 
where the first component $F_0$ is the set of values specified by $(\fname=time,\, \ftype=\texttt{datetime},\, \fttype=\reals)$ and $F_1$ is specified by $(\fname=temperature,\, \ftype=\texttt{float},\, \fttype=\reals)$ and is the set of values $\fttype=\reals$. In the 3D fiber in \autoref{fig:fiber_example_cube}, time is replaced with location. This location variable is of type \texttt{point} and has two components \textit{latitude} and \textit{longitude} $\{(lat,lon) \in \reals^{2} \mid  -90\leq lat \leq 90,\, 0 \leq lon \leq 360\}$. The fiber for this dataset is
\begin{equation*}
    \dfiber = \reals \times \reals^{2} = \reals \times \reals \times \reals
\end{equation*} 
where the dimensionality of the fiber does not change, but the components of the fiber can be coupled. For example, \textit{location} can can either be specified as \((\fname = location,\, \ftype = \texttt{point},\, \fttype=\reals^{2})\)or \((\fname=latitude,\, \ftype=\texttt{float}, \, \fttype=\reals)\) and \((\fname=longitude,\, \ftype=\texttt{float},\,\fttype=\reals)\).

@@

\subsubsection{Measurement Scales: Monoid Actions}
\label{sec:data_monoid}
Implementing expressive visual encodings requires formally describing the structure on the components of the fiber, which we define by the actions of a monoid on the component. In doing so, we specify the properties of the component that must be preserved in a graphic representation. A monoid \cite{Monoid2021} $\monoid$ is a set with a binary operation $\ast:\monoid \times \monoid\rightarrow \monoid$ that satisfies the axioms:
\begin{align*}
    \textbf{associativity}\; & \text{for all} a, b, c \in \monoid \, (a \bullet b) \bullet c = a \bullet (b \bullet c)\\
    \textbf{identity}\; & \text{for all } a\in \monoid,\,  e\bullet a = a 
\end{align*} 
As defined on a component of \dfiber, a left monoid action \cite{SemigroupAction2021,nlab:action} of $\monoid_i$ is a set $\dfiber_i$ with an action $\bullet: \monoid\times \dfiber_i \rightarrow \dfiber_i$ with the properties:
\begin{align*}
    \textbf{associativity}\;& \text{for all } f,g \in \monoid_i \text{ and } x\in \dfiber_i,\, f\bullet(g\bullet x) = (f\ast g) \bullet x\\
    \textbf{identity}\;& \text{for all } x\in \dfiber_i, e\in \monoid_i,\,  e\bullet x = x 
\end{align*}
The identity and associativity properties of the action denote that the action is a monoid homomorphism, which means that the group operation is preserved on both sides of the action\cite{weissteinGroupHomomorphism}.

\begin{figure}[H]
    \begin{subfigure}{.5\textwidth}
        \includegraphics[width=\textwidth]{figures/math/monoid_total.png}
        \caption{}
        \label{fig:monoid:total}
    \end{subfigure}
    \begin{subfigure}{.5\textwidth}
        \includegraphics[width=1\textwidth]{figures/math/monoid_partial.png}
        \caption{}
        \label{fig:monoid:partial}
    \end{subfigure}
    \caption{The action $\bullet$ in \autoref{fig:monoid:total} is the arrows from the partial order diagram of weather states on the left to the diagram of weather states on the right. Since the action maps the weather states to themselves, the ordering defined by the monoid $\ast$ is preserved on both sides of the action. The action in \autoref{fig:monoid:partial} is monoid homomorphism because the ordering of the weather states is the same as the ordering of the elements they are mapped to. Given $sun \geq clouds\geq rain\, lightining$ on the right, the action  $sun\, clouds \rightarrow clouds$, and $rain\, lightining \rightarrow rain$ is structure preserving because on the left $cloud \geq rain$ so the relative ordering of elements is the same as the elements they are mapped to.}
    \label{fig:monoid}
\end{figure}

One example of monoids are partial orderings on a set, such as seen in \label{fig:monoid}. Each hasse diagram of the set of weather states describes an ordering on the set; the arrow goes from the lesser value to the greater one. For example, $storm \leq rain$. In \autoref{fig:monoid:total}, the action $\bullet$ maps the elements of a set of weather states into itself by mapping them into other elements of the weather states. The action in  \autoref{fig:monoid:total}, represented as the arrows between the hasse diagrams of the weather states, maps the weather states to themselves; therefore the ordering of the weather states is identical on both sides of the action and it is therefore homomorphic. The action $\bullet$ in \autoref{fig:monoid:partial} is a monotone map\cite{fongInvitationAppliedCategory2019}, meaning that for  $a,b \in \dfiber_{i}$
\begin{equation}
    if\, a \leq b \; then \; \bullet(a) \leq \bullet(b)
\end{equation}
the structure the action must preserve is the relative, rather than exact, ordering. Since groups are monoids with invertible operations, this definition of structure is also broad enough to include the Steven's measurment scales\cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}. Monoids are also commonly found in functional programming because a property of monoids is that the components can be composed into complex transformation pipelines \cite{yorgeyMonoidsThemeVariations}.

As with the fiber \dfiber\, the total monoid space \monoid\ is the cartesian product
\begin{equation}
\monoid= \monoid_{0} \times \ldots \times \monoid_{i}\times \ldots \times\ldots \monoid_{n}
\end{equation}
of each monoid $\monoid_{i}$ on $\dfiber_{i}$.  The monoid is also added to the specification of the fiber $(\fname_i,\, \ftype_i,\, \fttype\, \monoid_i)$
